---
title: "Lab 04 - La Quinta is Spanish for next to Denny's, Pt. 1"
author: "Cynthia Jiao"
date: "02/03/2025"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dplyr)
library(dsbox) 
```

```{r load-data, message = FALSE}
states <- read_csv("data/states.csv")
```

### Exercise 1

In total, there are 1643 Denny's restaurants (each row in the tibble represents one Denny's location), and there are 6 variables (6 columns) that describes the address, city, state, zip code, longitude and latitude of each Denny's.

```{r exercise 1}
## two dataset are called dennys and laquinta
## dennys
print(dennys)
nrow(dennys)
ncol(dennys)


```


### Exercise 2

In total, there are 909 La quinta hotels (each row in tibble represents one La quinta locatin), and there are 6 variables (6 columns) that describes the address, city, state, zip code, longitude and latitude of each La quinta.

```{r exercise 2}
## laquinta
print(laquinta)
nrow(laquinta)
ncol(laquinta)

```

### Exercise 3

There are multiple locations of La quinta that are outside of the U.S. In total, there are 25 locations outside of the U.S. (If I am counting correctly), including locations in Canada, Mexico, China, New Zealand, Turkey, United Arab Emirates, Chile, Colombia, and Ecuador. 

As for Denny's, their website did not specify which or the total number of locations that are outside of the U.S. However, they do have locations in the following countries: Canada, Puerto Rico, Guam, Mexico, Honduras, Costa Rica, Guatemala, El Salvador, The Philippines, Indonesia, New Zealand, The United Arab Emirates, The United Kingdom, and Japan.

### Exercise 4

One way to filter out the non-US locations would be using US state abbreviations. Use dplyr and enter all 50 US state abbreviations to filter out all observations that have non-US state abbreviations. For example, there is one La quinta in Colombia with state column saying "ANT," which will be filtered out if using this method.

The other way to do it is through US zip code range. According to ChatGPT, all US zip codes fall within the range of 00501 to 99950. Although not every consecutive number within the range is assigned as an existing US zip code, we can still use dplyr to filter out any location with zip code outside of this range. Or, all existing US zip code can be scraped from USPS website, and a filter that contains all existing zip codes can be applied to the dennys and laquinta data to filter out any non-US locations.

### Exercise 5

No, there is not any Denny's location outside of the US in the dataset.

```{r exercise 5}
################### dn and lq means dennys and laquinta dataset, correct???

## finding any non-US Denny's location
dennys %>%
  filter(!(state %in% states$abbreviation))


```


### Exercise 6

```{r exercise 6}
## creating country variable for dennys
dennys %>%
  mutate(country = "United States")

```

### Exercise 7

Yes, there are 14 La quinta locations outside of the US. 

10 locations in Mexico;
2 locations in Canada;
1 location in Colombia;
1 location in Honduras.

```{r exercise 7}

## finding any non-US La quinta locations
laquinta %>%
  filter(!(state %in% states$abbreviation))


```


### Exercise 8

```{r exercise 8}

## adding a country variable to the La Quinta dataset AND filtering out all non-US La Quinta locations

laquinta %>%
  mutate(country = case_when(
    state %in% state.abb ~ "United States",
    state %in% c("ON", "BC") ~ "Canada",
    state == "ANT" ~ "Colombia",
    state == "FM" ~ "Honduras",
    state %in% c("AG", "QR", "CH", "NL", "VE","PU","Sl") ~ "Mexico"
  )) %>%
  filter(country == "United States")


```

### Exercise 9

California has the most Denny's locations (403 locations), and Delaware has least Denny's (just 1 location). I am not surprised about California being the state that has the most Denny's locations, as Denny's originated from California. But I am not sure why Delaware has the least location, maybe they already have similar local brand that is more popular than Denny's?

Texas has the most La quinta locations (237 locations), but DC, Delaware, Hawaii have no La quinta hotel. Similar to Denny's, I am not surprised about Texas being the state that has the most La quinta locations, as they originated from Texas. But I don't know why there are 3 states without any La quinta locations, maybe the competition is too fierce at these places??
```{r exercise 9}

## counting the states have the most and fewest Denny’s locations

dennys %>%
  count(state) %>%
  inner_join(states, by = c("state" = "abbreviation"))

## counting the states have the most and fewest La quinta’s locations

laquinta %>%
  count(state) %>%
  inner_join(states, by = c("state" = "abbreviation"))

```

### Exercise 10

DC has the most Denny's locations per thousand square miles. For every 1000 square miles, DC has 29 locations of Denny's. Since DC only has 2 Denny's, this high density is probably because DC is the smallest state.

Rhode Island has the most La quinta locations per thousand square miles. For every 1000 square miles, RI has 1 location of La quinta.

```{r exercise 10}

## counting states having the most Denny's location per thousand square miles.

dennys_density <- dennys %>%
  count(state) %>%
  inner_join(states, by = c("state" = "abbreviation")) %>%
  mutate(dennys_per_1000_sq_miles = n / (area / 1000)) %>%
  arrange(desc(dennys_per_1000_sq_miles))

print(dennys_density)

## counting states having the most La quinta location per thousand square miles.

lq_density <- laquinta %>%
  count(state) %>%
  inner_join(states, by = c("state" = "abbreviation")) %>%
  mutate(lq_per_1000_sq_miles = n / (area / 1000)) %>%
  arrange(desc(lq_per_1000_sq_miles))

print(lq_density)

```

### Exercise 11

Just from the NC scatter plot, Mitch Hedberg’s joke does not hold true. Out of 40 locations of Denny's and La quinta in NC, only 7 of them (3 pairs) seem to be next to each other. 

```{r exercise 11-1}

## combining two dataset
dennys <- dennys %>%
  mutate(establishment = "Denny's")

laquinta <- laquinta %>%
  mutate(establishment = "La Quinta")

dn_lq <- bind_rows(dennys, laquinta)

## scatter plot for ALL the ds and lq locations

ggplot(dn_lq, mapping = aes(
  x = longitude,
  y = latitude,
  color = establishment
)) +
  geom_point()

```
```{r exercise 11-2}

## filter the data for observations in North Carolina only

dn_lq_nc <- dn_lq %>%
  filter(state == "NC")

## recreate the scatter plot for NC

ggplot(dn_lq_nc, mapping = aes(
  x = longitude,
  y = latitude,
  color = establishment
)) +
  geom_point(alpha = 0.7)


```

### Exercise 12

From Texas scatter plot, Mitch Hedberg’s joke holds true. Although I cannot count the exact numbers of Denny's and La quinta that are next to each other, there are clearly many of them clustering together on the graph. 

```{r exercise 12}

## filter the data for observations in Texas only

dn_lq_tx <- dn_lq %>%
  filter(state == "TX")

## recreate the scatter plot for TX

ggplot(dn_lq_tx, mapping = aes(
  x = longitude,
  y = latitude,
  color = establishment
)) +
  geom_point(alpha = 0.3)


```

